---
title: "Cardiovascular Disease - Data Cleaning, EDA, Linear Model & Poisson GLM"
subtitle: 'Applied Machine Learning and Predictive Modelling 1 - FS26'
author: "Hanieh Jebeli, Enerel Khuyag, Aurelio Wyrsch"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.width  = 8,
  fig.height = 5
)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(GGally)
library(knitr)
library(broom)
library(gridExtra)
library(scales)
```

---

# 1. Data Source & Context

*Section lead: Hanieh Jebeli*

The dataset comes from Kaggle (Sulianova, 2019) and contains medical records
for **70,000 patients**, each described by 11 clinical features and a binary
outcome indicating the presence or absence of cardiovascular disease (`cardio`).
Data was collected during medical examinations and covers demographic information,
blood pressure readings, blood test results, and self-reported lifestyle habits.

**Research question:**

> *"Which patient characteristics are most associated with cardiovascular disease,
> and can we predict its presence from routine clinical measurements?"*

---

# 2. Data Loading & Initial Inspection


```{r load-data}
cardio_raw <- read.csv("../data/cardio_train.csv", sep = ";")

dim(cardio_raw)
head(cardio_raw, 5)
str(cardio_raw)
```

The raw dataset has **`r nrow(cardio_raw)` rows** and **`r ncol(cardio_raw)` columns**.

| Variable | Type | Description |
|---|---|---|
| `id` | Integer | Patient ID — dropped in analysis |
| `age` | Integer | Age in **days** — converted to years |
| `gender` | Binary | 1 = Female, 2 = Male |
| `height` | Integer | Height in cm |
| `weight` | Float | Weight in kg |
| `ap_hi` | Integer | Systolic blood pressure (mmHg) |
| `ap_lo` | Integer | Diastolic blood pressure (mmHg) |
| `cholesterol` | Categorical | 1 = Normal, 2 = Above normal, 3 = Well above normal |
| `gluc` | Categorical | 1 = Normal, 2 = Above normal, 3 = Well above normal |
| `smoke` | Binary | 0 = No, 1 = Yes |
| `alco` | Binary | 0 = No, 1 = Yes |
| `active` | Binary | 0 = No, 1 = Yes |
| `cardio` | Binary | **Target** — 0 = No disease, 1 = Disease |

---

# 3. Data Cleaning



## 3.1 Recode and engineer variables

```{r recode}
cardio <- cardio_raw %>%
  select(-id) %>%
  mutate(
    # Age: days -> years
    age_years = round(age / 365.25, 1),

    # Body Mass Index
    bmi = weight / (height / 100)^2,

    # Pulse pressure
    pulse_pressure = ap_hi - ap_lo,

    # Gender as factor
    gender = factor(gender, levels = c(1, 2),
                    labels = c("Female", "Male")),

    # Cholesterol — 3 ordered levels (satisfies course requirement)
    cholesterol = factor(cholesterol, levels = c(1, 2, 3),
                         labels = c("Normal", "Above normal", "Well above normal"),
                         ordered = TRUE),

    # Glucose — 3 ordered levels
    gluc = factor(gluc, levels = c(1, 2, 3),
                  labels = c("Normal", "Above normal", "Well above normal"),
                  ordered = TRUE),

    smoke  = factor(smoke,  levels = c(0, 1), labels = c("No", "Yes")),
    alco   = factor(alco,   levels = c(0, 1), labels = c("No", "Yes")),
    active = factor(active, levels = c(0, 1), labels = c("No", "Yes")),
    cardio = factor(cardio, levels = c(0, 1), labels = c("No disease", "Disease"))
  ) %>%
  select(-age)

glimpse(cardio)
```

## 3.2 Missing values

```{r missing}
colSums(is.na(cardio))
```

No missing values across all variables.

## 3.3 Remove physiologically impossible values

Inspection of the raw data reveals clear entry errors in blood pressure,
height, and weight. These are medically impossible values that would distort
every model if kept.

| Problem | Count |
|---|---|
| Systolic BP outside 60-250 mmHg | 228 |
| Diastolic BP outside 40-150 mmHg | 1,034 |
| Systolic BP <= diastolic BP | 1,236 |
| Height outside 140-210 cm | 153 |
| Weight outside 30-200 kg | 7 |

After applying all filters simultaneously (rows can fail multiple conditions):

```{r clean}
cat("Rows before cleaning:", nrow(cardio), "\n")

cardio_clean <- cardio %>%
  filter(
    ap_hi  >= 60,  ap_hi  <= 250,
    ap_lo  >= 40,  ap_lo  <= 150,
    ap_hi  >  ap_lo,
    height >= 140, height <= 210,
    weight >= 30,  weight <= 200,
    bmi    >= 10,  bmi    <= 60
  )

cat("Rows after cleaning:", nrow(cardio_clean), "\n")
cat("Rows removed:       ", nrow(cardio) - nrow(cardio_clean), "\n")
cat("Percentage removed: ", round((nrow(cardio) - nrow(cardio_clean)) / nrow(cardio) * 100, 2), "%\n")
```

We removed **`r nrow(cardio) - nrow(cardio_clean)`** rows
(`r round((nrow(cardio) - nrow(cardio_clean))/nrow(cardio)*100, 1)`% of the data),
leaving **`r nrow(cardio_clean)`** clean patient records.

## 3.4 Dataset summary

```{r summary}
summary(cardio_clean %>%
          select(age_years, height, weight, bmi, ap_hi, ap_lo, pulse_pressure))
```

```{r target-balance}
cardio_clean %>%
  count(cardio) %>%
  mutate(pct = paste0(round(100 * n / sum(n), 1), "%")) %>%
  kable(caption = "Target variable distribution")
```

Classes are almost perfectly balanced (50/50) — no resampling required for modelling.

## 3.5 Save cleaned data for teammates

```{r save}
saveRDS(cardio_clean, "../data/cardio_clean.rds")
cat("Saved: ../data/cardio_clean.rds\n")
cat("Enerel and Aurelio: load with readRDS('../data/cardio_clean.rds')\n")
```

---

