---
title: "Bank Marketing - Data Cleaning, EDA, Linear Model & Poisson GLM"
subtitle: 'Applied Machine Learning and Predictive Modelling 1 - FS26'
author: "Hanieh Jebeli (lead), Enerel Khuyag, Aurelio Wyrsch"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.width  = 8,
  fig.height = 5
)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(GGally)
library(knitr)
library(broom)
library(gridExtra)
library(scales)
```

---

# 1. Data Source & Context

*Section lead: Hanieh Jebeli*

The dataset comes from the UCI Machine Learning Repository and contains records
of **45,211 phone marketing calls** made by a Portuguese bank between May 2008
and November 2010. Each row represents one call to one client, with 16 features
describing the client's profile, the campaign contact history, and the outcome
of previous campaigns.

**Research question:**

> *"Can we predict which bank customers will subscribe to a term deposit
> before we call them — and what drives their decision?"*

This is a practical business problem: a model that correctly identifies likely
subscribers allows a bank to focus its marketing resources on the right people,
reducing cost and improving conversion rates.

---

# 2. Data Loading & Initial Inspection

*Section lead: Hanieh Jebeli*

```{r load-data}
# File uses semicolons as separator
bank_raw <- read.csv("../data/bank-full.csv", sep = ";")

dim(bank_raw)
head(bank_raw, 5)
str(bank_raw)
```

The raw dataset has **`r nrow(bank_raw)` rows** and **`r ncol(bank_raw)` columns**.

| Variable   | Type        | Description                                                   |
|:-----------|:------------|:--------------------------------------------------------------|
| `age`      | Continuous  | Client age in years                                           |
| `job`      | Categorical | Type of job (12 levels)                                       |
| `marital`  | Categorical | Marital status (married, single, divorced)                    |
| `education`| Categorical | Education level (primary, secondary, tertiary, unknown)       |
| `default`  | Binary      | Has credit in default? (yes/no)                               |
| `balance`  | Continuous  | Average yearly balance in euros                               |
| `housing`  | Binary      | Has housing loan? (yes/no)                                    |
| `loan`     | Binary      | Has personal loan? (yes/no)                                   |
| `contact`  | Categorical | Contact type (cellular, telephone, unknown)                   |
| `day`      | Continuous  | Last contact day of month                                     |
| `month`    | Categorical | Last contact month (jan-dec)                                  |
| `duration` | Continuous  | Last call duration in seconds                                 |
| `campaign` | **Count**   | Number of contacts during this campaign                       |
| `pdays`    | Continuous  | Days since last contact (-1 = never contacted before)        |
| `previous` | **Count**   | Number of contacts before this campaign                       |
| `poutcome` | Categorical | Outcome of previous campaign (success, failure, other, unknown)|
| `y`        | Binary      | **Target** — subscribed? (yes/no)                            |

---

# 3. Data Cleaning

*Section lead: Hanieh Jebeli*

## 3.1 Recode and engineer variables

```{r recode}
bank <- bank_raw %>%
  mutate(
    # Target: yes/no -> factor
    subscribed = factor(y, levels = c("no", "yes"),
                        labels = c("No", "Yes")),

    # Job: 12 levels — satisfies course 3+ categorical requirement
    job = factor(job),

    # Marital: 3 levels
    marital = factor(marital,
                     levels = c("single", "married", "divorced")),

    # Education: 4 ordered levels
    education = factor(education,
                       levels = c("unknown", "primary", "secondary", "tertiary"),
                       ordered = TRUE),

    # Binary variables as proper factors
    default = factor(default, levels = c("no", "yes")),
    housing = factor(housing, levels = c("no", "yes")),
    loan    = factor(loan,    levels = c("no", "yes")),

    # Contact type
    contact = factor(contact),

    # Month as ordered factor
    month = factor(month,
                   levels = c("jan","feb","mar","apr","may","jun",
                              "jul","aug","sep","oct","nov","dec"),
                   ordered = TRUE),

    # Previous campaign outcome
    poutcome = factor(poutcome,
                      levels = c("unknown","other","failure","success")),

    # pdays: -1 means never contacted — recode to NA for clarity
    pdays_clean = ifelse(pdays == -1, NA, pdays),

    # Flag: was client contacted before this campaign?
    prev_contacted = factor(ifelse(previous == 0, "No", "Yes")),

    # log balance — balance is an amount variable, needs log transform for LM
    # some balances are negative (overdraft) — shift first
    balance_shifted = balance - min(balance) + 1,
    log_balance     = log(balance_shifted)
  ) %>%
  select(-y)   # drop original target — we use 'subscribed' instead

glimpse(bank)
```

## 3.2 Missing values

```{r missing}
colSums(is.na(bank))
```

No missing values in the original columns. The `pdays_clean` column has NAs by
design — these represent clients who were never contacted before this campaign
(coded as -1 in the raw data). This is expected and handled correctly.

## 3.3 Check for outliers

```{r outliers}
# Campaign: number of contacts — check for unreasonable values
cat("campaign range:", range(bank$campaign), "\n")
cat("campaign > 30: ", sum(bank$campaign > 30), "rows\n")

# Duration: call length in seconds
cat("duration range:", range(bank$duration), "\n")
cat("duration = 0:  ", sum(bank$duration == 0), "rows\n")

# Balance: can be negative (overdraft) — that is valid
cat("balance range:", range(bank$balance), "\n")
```

```{r remove-outliers}
cat("Rows before cleaning:", nrow(bank), "\n")

bank_clean <- bank %>%
  filter(
    campaign <= 30,    # more than 30 calls to one client is a data error
    duration > 0       # duration = 0 means call never happened — remove
  )

cat("Rows after cleaning: ", nrow(bank_clean), "\n")
cat("Rows removed:        ", nrow(bank) - nrow(bank_clean), "\n")
```

## 3.4 Dataset summary

```{r summary}
summary(bank_clean %>%
          select(age, balance, duration, campaign, previous))
```

```{r target-balance}
bank_clean %>%
  count(subscribed) %>%
  mutate(pct = paste0(round(100 * n / sum(n), 1), "%")) %>%
  kable(caption = "Target variable distribution")
```

The classes are **imbalanced** — around 88% did not subscribe and 12% did.
This is realistic for marketing data and will need to be considered when
evaluating model performance (accuracy alone is not enough — we will use AUC).

## 3.5 Save cleaned data for teammates

```{r save}
saveRDS(bank_clean, "../data/bank_clean.rds")
cat("Saved: ../data/bank_clean.rds\n")
cat("Enerel and Aurelio: load with readRDS('../data/bank_clean.rds')\n")
```

---

# 4. Exploratory Data Analysis

*Section lead: Hanieh Jebeli*

## 4.1 Subscription rate by job type

```{r eda-job}
bank_clean %>%
  group_by(job) %>%
  summarise(rate = mean(subscribed == "Yes")) %>%
  arrange(desc(rate)) %>%
  ggplot(aes(x = reorder(job, rate), y = rate, fill = rate)) +
  geom_col(alpha = 0.85) +
  scale_fill_gradient(low = "steelblue", high = "tomato") +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  labs(title = "Subscription rate by job type",
       x = "", y = "Subscription rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

Students and retired clients have the highest subscription rates. Blue-collar
and services workers show the lowest rates — likely reflecting different
financial priorities and risk appetites.

## 4.2 Age and balance distribution by outcome

```{r eda-age-balance}
p1 <- ggplot(bank_clean, aes(x = subscribed, y = age, fill = subscribed)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Age by subscription", x = "", y = "Age") +
  theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(bank_clean, aes(x = subscribed, y = balance, fill = subscribed)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.3) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Balance by subscription", x = "", y = "Balance (euros)") +
  theme_minimal() + theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

Subscribers tend to be slightly older and have higher account balances,
suggesting financial stability is positively associated with subscription.

## 4.3 Call duration — strongest single predictor

```{r eda-duration}
ggplot(bank_clean, aes(x = duration / 60, fill = subscribed)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Call duration by subscription outcome",
       subtitle = "Longer calls strongly associated with subscription",
       x = "Duration (minutes)", y = "Count", fill = "Subscribed") +
  theme_minimal()
```

Call duration is the single strongest predictor in the dataset. Customers
who subscribed had much longer calls on average. Note: this variable cannot
be used for prediction in a real deployment (we don't know duration before
calling) — but it is valid for a retrospective analysis.

## 4.4 Campaign contacts — count variable

```{r eda-campaign}
bank_clean %>%
  filter(campaign <= 15) %>%
  ggplot(aes(x = campaign, fill = subscribed)) +
  geom_bar(position = "fill", alpha = 0.85) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = percent) +
  labs(title = "Subscription rate by number of campaign contacts",
       x = "Number of calls (campaign)", y = "Proportion",
       fill = "Subscribed") +
  theme_minimal()
```

Subscription rate drops sharply after 2-3 contacts. Clients contacted many
times are much less likely to subscribe — diminishing returns are clearly visible.

## 4.5 Smoother: age vs call duration

```{r eda-smoother}
ggplot(bank_clean, aes(x = age, y = duration / 60, color = subscribed)) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(title = "Call duration vs age (LOESS smoother)",
       x = "Age (years)", y = "Duration (minutes)", color = "Subscribed") +
  theme_minimal()
```

## 4.6 Correlation matrix

```{r eda-corr}
bank_clean %>%
  select(age, balance, duration, campaign, previous, day) %>%
  cor(use = "complete.obs") %>%
  round(2) %>%
  kable(caption = "Pearson correlations — continuous and count variables")
```

Most variables show weak mutual correlations, which is good — it means
multicollinearity is not a major concern for the models.

---

# 5. Feature Engineering

*Section lead: Hanieh Jebeli*

```{r features}
bank_model <- bank_clean %>%
  mutate(
    # Numeric target for aggregation (Poisson)
    subscribed_num = as.integer(subscribed == "Yes"),

    # Age groups for Poisson aggregation
    age_group = cut(age,
                    breaks = c(0, 30, 40, 50, 60, 100),
                    labels = c("<30", "30-40", "40-50", "50-60", "60+"),
                    right  = FALSE),

    # Duration in minutes — more interpretable
    duration_min = duration / 60,

    # Education as numeric for LM
    education_num = as.integer(education)
  )

glimpse(bank_model %>% select(subscribed_num, age_group, duration_min, education_num))
```

---

