---
title: "Bank Marketing - Binomial GLM & Generalised Additive Model"
subtitle: 'Applied Machine Learning and Predictive Modelling 1 - FS26'
author: "Hanieh Jebeli, Enerel Khuyag (lead), Aurelio Wyrsch"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = FALSE,
  fig.width  = 8,
  fig.height = 5
)
options(yaml.eval.expr = FALSE)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(mgcv)
library(pROC)
library(caret)
library(broom)
library(gridExtra)
library(scales)
library(knitr)
```

---

# 1. Load

```{r load}
bank_clean <- readRDS("~/bank-marketing-ml/data/bank_clean.rds")

bank_clean <- bank_clean %>%
  mutate(subscribed_bin = as.integer(subscribed == "Yes"))

glimpse(bank_clean)
```

---

# 2. Binomial GLM - Predicting subscription

*Section lead: Enerel Khuyag*

## 2.1 Motivation

Since the response variable **subscribed** is binary (Yes/No), a **Binomial GLM
with a logit link** is the natural modelling choice. Unlike a plain linear model,
it constrains predicted values to the (0, 1) probability range and models the
**log-odds** of subscription as a linear combination of predictors. This makes
the coefficients directly interpretable as odds ratios after exponentiation.

## 2.2 Train/Test Split

To evaluate the model on unseen data we use the same 80/20 stratified split
strategy as Aurelio's section, ensuring comparable performance metrics across all models.

```{r glm-split}
set.seed(123)
indices <- createDataPartition(bank_clean$subscribed, p = 0.80, list = FALSE)

bank_train <- bank_clean %>% slice(indices)
bank_test  <- bank_clean %>% slice(-indices)

bank_test_truth     <- bank_test$subscribed
bank_test_truth_bin <- bank_test$subscribed_bin

cat("Training rows:", nrow(bank_train), "\n")
cat("Test rows:    ", nrow(bank_test),  "\n")
cat("Subscription rate (train):", round(mean(bank_train$subscribed == "Yes"), 3), "\n")
cat("Subscription rate (test): ", round(mean(bank_test$subscribed  == "Yes"), 3), "\n")
```

## 2.3 Model Fitting

```{r glm-fit}
glm_fit <- glm(
  subscribed_bin ~ age + job + marital + education +
                   default + balance + housing + loan +
                   contact + month + campaign +
                   previous + poutcome,
  data   = bank_train,
  family = binomial(link = "logit")
)

saveRDS(glm_fit, "~/bank-marketing-ml/models/binomial_fit.rds")

summary(glm_fit)
```

## 2.4 Coefficient Interpretation

```{r glm-coef}
tidy(glm_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)", p.value < 0.05) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(desc(estimate)) %>%
  kable(caption = "Significant odds ratios (p < 0.05) — Binomial GLM")
```

**Key interpretations:**

- **`poutcome = success`**: The strongest predictor by far. Clients whose
  previous campaign ended in a subscription are roughly 8–10× more likely to
  subscribe again — past behaviour is the best predictor of future behaviour.

- **`contact = cellular`**: Clients contacted by mobile phone have higher
  subscription odds than those reached by landline or unknown contact type.

- **`month` effects**: Calls in March, September, October, and December show
  higher subscription rates, likely reflecting seasonal financial planning cycles.

- **`housing = yes`**: Clients with a housing loan are less likely to subscribe —
  those already carrying mortgage debt may be unwilling to lock up additional capital.

- **`campaign`**: Each additional call during the current campaign slightly
  *reduces* the odds — repeated contact with uninterested clients is counterproductive.

```{r glm-coef-plot, fig.cap="Top 15 predictors by odds ratio (log scale)"}
tidy(glm_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(log(estimate)), n = 15) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate,
             ymin = conf.low, ymax = conf.high, colour = estimate > 1)) +
  geom_pointrange() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  coord_flip() +
  scale_y_log10() +
  scale_colour_manual(values = c("tomato", "steelblue"), guide = "none") +
  labs(title = "Odds Ratios — Binomial GLM (top 15 predictors)",
       x = NULL, y = "Odds Ratio (log scale)") +
  theme_minimal()
```

## 2.5 Model Evaluation

```{r glm-eval}
pred_prob_glm  <- as.vector(predict(glm_fit, newdata = bank_test, type = "response"))
pred_class_glm <- factor(if_else(pred_prob_glm > 0.5, "Yes", "No"),
                         levels = c("No", "Yes"))

cm_glm  <- confusionMatrix(pred_class_glm, bank_test_truth, positive = "Yes")
roc_glm <- roc(bank_test_truth_bin, pred_prob_glm)
auc_glm <- as.numeric(auc(roc_glm))

data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Balanced Accuracy", "AUC"),
  Value  = round(c(
    cm_glm$overall["Accuracy"],
    cm_glm$byClass["Sensitivity"],
    cm_glm$byClass["Specificity"],
    cm_glm$byClass["Balanced Accuracy"],
    auc_glm
  ), 3)
) %>% kable(caption = "Binomial GLM — Performance Metrics", row.names = FALSE)
```

```{r glm-confusion, fig.cap="Confusion matrix — Binomial GLM"}
as.data.frame(cm_glm$table) %>%
  mutate(label = case_when(
    Prediction == "No"  & Reference == "No"  ~ "True Negative",
    Prediction == "Yes" & Reference == "Yes" ~ "True Positive",
    Prediction == "Yes" & Reference == "No"  ~ "False Positive",
    Prediction == "No"  & Reference == "Yes" ~ "False Negative"
  )) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = label)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(label, "\nn = ", Freq)),
            size = 5, fontface = "bold", colour = "white") +
  scale_fill_manual(values = c(
    "True Positive"  = "#2ecc71", "True Negative"  = "#2ecc71",
    "False Positive" = "#e74c3c", "False Negative" = "#e74c3c"
  )) +
  labs(title = "Confusion Matrix — Binomial GLM", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r glm-roc, fig.cap="ROC curve — Binomial GLM"}
ggroc(roc_glm, colour = "steelblue", size = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.3, y = 0.1,
           label = paste0("AUC = ", round(auc_glm, 3)),
           colour = "steelblue", size = 5) +
  labs(title = "ROC Curve — Binomial GLM", x = "Specificity", y = "Sensitivity") +
  theme_minimal()
```

## 2.6 Discussion

The Binomial GLM provides a strong, interpretable baseline for predicting
subscription. The AUC indicates good discriminative power, and the coefficients
offer actionable business insights. The main limitation is the assumption of a
**linear relationship** between each continuous predictor and the log-odds,
which the GAM below relaxes.

---

# 3. Generalised Additive Model (GAM)

*Section lead: Enerel Khuyag*

## 3.1 Motivation

The Binomial GLM assumes that predictors like `age`, `balance`, and `campaign`
affect the log-odds **linearly**. A **GAM** replaces these linear terms with
flexible smooth spline functions fitted from the data, allowing the model to
capture non-linear relationships automatically.

## 3.2 Model Fitting

```{r gam-fit}
gam_fit <- gam(
  subscribed_bin ~ s(age) + s(balance) + s(campaign) + s(previous) +
                   job + marital + education +
                   default + housing + loan +
                   contact + month + poutcome,
  data   = bank_train,
  family = binomial(link = "logit"),
  method = "REML"
)

saveRDS(gam_fit, "~/bank-marketing-ml/models/gam_fit.rds")

summary(gam_fit)
```

## 3.3 Smooth Term Plots

```{r gam-plots, fig.height=8, fig.cap="Estimated smooth effects on log-odds of subscription"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(gam_fit, shade = TRUE, shade.col = "lightblue",
     seWithMean = TRUE, scale = 0, residuals = FALSE)
par(mfrow = c(1, 1))
```

**Interpretation of smooth terms:**

- **`s(age)`**: Non-linear U-shape — very young (~18–25) and older clients (~60+)
  are more likely to subscribe; middle-aged clients (40–55) are least likely.

- **`s(balance)`**: Higher balances positively predict subscription, flattening
  at very high values. Clients in overdraft are least likely to subscribe.

- **`s(campaign)`**: Subscription probability drops sharply after 2–3 contacts,
  confirming the diminishing returns pattern seen in the EDA.

- **`s(previous)`**: Clients with 1–3 prior contacts show higher subscription rates.

## 3.4 Model Evaluation

```{r gam-eval}
pred_prob_gam  <- as.vector(predict(gam_fit, newdata = bank_test, type = "response"))
pred_class_gam <- factor(if_else(pred_prob_gam > 0.5, "Yes", "No"),
                         levels = c("No", "Yes"))

cm_gam  <- confusionMatrix(pred_class_gam, bank_test_truth, positive = "Yes")
roc_gam <- roc(bank_test_truth_bin, pred_prob_gam)
auc_gam <- as.numeric(auc(roc_gam))

data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Balanced Accuracy", "AUC"),
  Value  = round(c(
    cm_gam$overall["Accuracy"],
    cm_gam$byClass["Sensitivity"],
    cm_gam$byClass["Specificity"],
    cm_gam$byClass["Balanced Accuracy"],
    auc_gam
  ), 3)
) %>% kable(caption = "GAM — Performance Metrics", row.names = FALSE)
```

```{r gam-confusion, fig.cap="Confusion matrix — GAM"}
as.data.frame(cm_gam$table) %>%
  mutate(label = case_when(
    Prediction == "No"  & Reference == "No"  ~ "True Negative",
    Prediction == "Yes" & Reference == "Yes" ~ "True Positive",
    Prediction == "Yes" & Reference == "No"  ~ "False Positive",
    Prediction == "No"  & Reference == "Yes" ~ "False Negative"
  )) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = label)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(label, "\nn = ", Freq)),
            size = 5, fontface = "bold", colour = "white") +
  scale_fill_manual(values = c(
    "True Positive"  = "#2ecc71", "True Negative"  = "#2ecc71",
    "False Positive" = "#e74c3c", "False Negative" = "#e74c3c"
  )) +
  labs(title = "Confusion Matrix — GAM", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## 3.5 GLM vs GAM Comparison

```{r compare-roc, fig.cap="ROC curves: Binomial GLM vs GAM"}
ggroc(list(GLM = roc_glm, GAM = roc_gam), size = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c(GLM = "steelblue", GAM = "tomato")) +
  annotate("text", x = 0.25, y = 0.12,
           label = paste0("AUC GLM = ", round(auc_glm, 3)),
           colour = "steelblue", size = 4) +
  annotate("text", x = 0.25, y = 0.06,
           label = paste0("AUC GAM = ", round(auc_gam, 3)),
           colour = "tomato", size = 4) +
  labs(title = "ROC Curves: Binomial GLM vs GAM",
       x = "Specificity", y = "Sensitivity", colour = "Model") +
  theme_minimal()
```

```{r compare-table}
tibble(
  Model               = c("Binomial GLM", "GAM"),
  AUC                 = c(round(auc_glm, 4), round(auc_gam, 4)),
  Accuracy            = c(round(cm_glm$overall["Accuracy"], 4),          round(cm_gam$overall["Accuracy"], 4)),
  Sensitivity         = c(round(cm_glm$byClass["Sensitivity"], 4),       round(cm_gam$byClass["Sensitivity"], 4)),
  Specificity         = c(round(cm_glm$byClass["Specificity"], 4),       round(cm_gam$byClass["Specificity"], 4)),
  `Balanced Accuracy` = c(round(cm_glm$byClass["Balanced Accuracy"], 4), round(cm_gam$byClass["Balanced Accuracy"], 4))
) %>% kable(caption = "Binomial GLM vs GAM — side-by-side comparison")
```

## 3.6 Discussion

The GAM captures non-linear effects the Binomial GLM misses — particularly the
U-shaped age effect and diminishing returns for campaign contacts. For **model
transparency**, the GLM remains valuable. For **prediction accuracy**, the GAM
is preferable. Both models confirm the same core insight: focus on clients with
a successful previous campaign outcome, contacted by mobile, in their first 1–2
calls of the current campaign.

---

# 4. Session Info

```{r session-info}
sessionInfo()
```
