---
title: "Bank Marketing - Cross Validation, Support Vector Machine and Neural Networks"
subtitle: 'Applied Machine Learning and Predictive Modelling 1 - FS26'
author: "Hanieh Jebeli, Enerel Khuyag, Aurelio Wyrsch (lead)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.width  = 8,
  fig.height = 5
)
options(yaml.eval.expr = FALSE)
knitr::opts_knit$set(aliases = NULL)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(GGally)
library(knitr)
library(broom)
library(gridExtra)
library(scales)
library(e1071)
library(caret)
library(nnet)
```

---

# 1. Load

```{r load, include=FALSE}
bank_clean <- readRDS("../data/bank_clean.rds") %>% 
  select(-duration, -pdays_clean, -prev_contacted, -balance_shifted, -log_balance)
```

---

# 2. Train/Test Split

*Section lead: Aurelio Wyrsch*

To evaluate how well a model generalizes to unseen data, we split ***bank_clean** into a training set (80%) and a test set (20%). Since the target variable **subscribed** is heavily imbalanced — 88.3% "No" vs 11.7% "Yes" — a simple random split risks under-representing subscribers in one of the sets. We therefore use stratified sampling via createDataPartition(), which preserves the class proportions of subscribed in both the training and test set, ensuring neither set is accidentally dominated by one class.

```{r train-test-split}
set.seed(123)
indices <- createDataPartition(bank_clean$subscribed, p = .80, list = F)

bank_train <- bank_clean %>%
  slice(indices)

bank_test <- bank_clean %>%
  slice(-indices) %>% 
  select(-subscribed)

bank_test_truth <- bank_clean %>%
  slice(-indices) %>%
  pull(subscribed)
```

---

# 2. Support Vector Machine (SVM) - Predicting subscribed

...

**Model:** svm(subscribed ~ . , data = bank_train, kernel = "radial", cost = 10)

# 2.1. Cross Validation

```{r cv-svm}
ctrl_svm <- trainControl(
  method = "cv",      
  number = 3,     
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE,
  verboseIter     = TRUE 
)
```

# 2.2. Tune

```{r tune-svm}
tune_grid <- expand.grid(
  C     = c(0.01, 0.1, 1),
  sigma = c(0.001, 0.01, 0.1)
)

set.seed(123)
small_indices <- createDataPartition(bank_train$subscribed, p = 0., list = F)
bank_train_small <- bank_train %>% slice(small_indices)

set.seed(123)
bank_svm <- train(
  subscribed ~ .,
  data       = bank_train,
  method     = "svmRadial",
  trControl  = ctrl_svm,
  tuneGrid   = tune_grid,
  metric     = "ROC",
  preProcess = c("center", "scale"),
  weights    = ifelse(bank_train$subscribed == "Yes", 7.5, 1)
)

bank_svm$results

plot(bank_svm)
```

## 2.3 Evalute

```{r evaluate-svm}
pred_svm <- factor(as.character(predict(bank_svm, bank_test)),
                   levels = c("No", "Yes"))

cm_svm <- confusionMatrix(pred_svm, bank_test_truth)
cm_svm
```

---

# 3. Neural Networks - Predicting subscribed

...

## 3.1 Train

```{r train-nnt}
set.seed(123)
bank_nnt <- nnet(subscribed ~ . , data = bank_train, size = 5, decay = 0.1, maxit = 200)
```

# 3.2 Test

```{r test-nnt}
bank_test_pred_nnt <- predict(bank_nnt, bank_test, type = "class")
```

# 3.3 Evaluate

```{r evaluate-nnt}
bank_test_pred_nnt <- factor(bank_test_pred_nnt, levels = c("No", "Yes"))

confusionMatrix(bank_test_pred_nnt, bank_truth)
```

---

# 4. Summary



