---
title: "Bank Marketing - Cross Validation, Support Vector Machine and Neural Networks"
subtitle: 'Applied Machine Learning and Predictive Modelling 1 - FS26'
author: "Hanieh Jebeli, Enerel Khuyag, Aurelio Wyrsch (lead)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r packages-03, include=FALSE}
library(tidyverse)
library(ggplot2)
library(GGally)
library(knitr)
library(broom)
library(gridExtra)
library(scales)
library(e1071)
library(caret)
library(pROC)
library(nnet)
```

```{r load-03, include=FALSE}
bank_clean <- readRDS("../data/bank_clean.rds")
```

---

# 11. Support Vector Machine (SVM) - Predicting subscribed

*Section lead: Aurelio Wyrsch*

Support Vector Machines are well-suited for binary classification tasks. Our target variable subscribed is binary (Yes/No), representing whether a client agreed to a term deposit during a phone campaign. By training an SVM to identify which clients are most likely to subscribe, the bank can prioritize its outreach towards high-probability leads — reducing campaign costs and improving conversion rates. Clients flagged as unlikely to subscribe can still be contacted if the budget allows, but resources are allocated more efficiently by targeting the most promising candidates first.

# 11.1 Feature Selection

Several variables are dropped before model training. duration is excluded since call duration is only known after a call is made, it cannot be used to predict subscription in a real deployment. The remaining dropped variables (pdays_clean, prev_contacted, balance_shifted, log_balance) are engineered features created for earlier modelling sections that are not needed.

```{r feature-selection,}
bank_clean <- bank_clean %>% 
  select(-duration, -pdays_clean, -prev_contacted, -balance_shifted, -log_balance)
```

## 11.2 Train/Test Split

To evaluate how well a model generalizes unseen data, we split **bank_clean** into a training set (80%) and a test set (20%). Since the target variable **subscribed** is heavily imbalanced — around 88% "No" vs 12% "Yes" — a simple random split risks under-representing subscribers in one of the sets. We therefore use stratified sampling via **createDataPartition()**, which preserves the class proportions of subscribed in both the training and test set, ensuring neither set is accidentally dominated by one class.

```{r train-test-split}
set.seed(123)
indices <- createDataPartition(bank_clean$subscribed, p = .80, list = F)

bank_train <- bank_clean %>%
  slice(indices)

bank_test <- bank_clean %>%
  slice(-indices) %>% 
  select(-subscribed)

bank_test_truth <- bank_clean %>%
  slice(-indices) %>%
  pull(subscribed)
```


## 11.3 Cross Validation

To evaluate the SVM reliably, we use a 5-fold cross validation. Rather than evaluating the model on a single validation set, 5-fold cross validation splits the data into 5 folds, iteratively training on 4 folds and validating on the remaining fold. This process repeats 5 times so that every observation is used for validation exactly once. The result is a more stable and unbiased estimate of model performance compared to a single train/validation split.

```{r cross-validation-svm}
ctrl_svm <- trainControl(
  method = "cv",      
  number = 5,     
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE
  # verboseIter     = TRUE 
)
```

## 11.4 Hyperparameter Tuning

To optimize the SVM's performance, we tune two hyperparameters using a grid search. The grid search tests 20 combinations across cost = (0.01, 0.05, 0.1, 0.5, 1) and gamma = (0.05, 0.1, 0.5, 1), training and evaluating a separate model for each combination.

```{r hyperparameter-tuning-svm}
tune_grid <- expand.grid(
  C     = c(0.01, 0.05, 0.1, 0.5, 1),
  sigma = c(0.05, 0.1, 0.5, 1) # gamma is called sigma in the caret package for svmRadial
)
```

## 11.5 Training

In the training, the model is evaluated using the 5-fold cross validation and grid search defined earlier. The combination that achieves the highest ROC is automatically selected as the final model. In total, 100 models are trained and compared in this process. Because of the class imbalance, a weighting was also added. Due to the computational demands of training a model 100 times on a large dataset, only a stratified 30% subsample of **bank_train** was used to speed up the process.

**Model:** svm(subscribed ~ . , data = bank_train_small, kernel = "radial", gamma = x, cost = y), 

- x = c(0.05, 0.1, 0.5, 1) 
- y = c(0.01, 0.05, 0.1, 0.5, 1)

Out of all the combinations, the highest ROC score was achieved with cost = 0.05 and gamma = 0.05 at 0.7259016.

```{r train-svm}
set.seed(123)
small_indices <- createDataPartition(bank_train$subscribed, p = 0.3, list = F)
bank_train_small <- bank_train %>% slice(small_indices)

set.seed(123)
bank_svm <- train(
  subscribed ~ .,
  data       = bank_train_small,
  method     = "svmRadial",
  trControl  = ctrl_svm,
  tuneGrid   = tune_grid,
  metric     = "ROC",
  preProcess = c("center", "scale"),
  weights = ifelse(bank_train_small$subscribed == "Yes", 7.5, 1)
)
```

```{r train-results-svm}
bank_svm$results %>% 
  filter(C == bank_svm$bestTune$C, sigma == bank_svm$bestTune$sigma)
plot(bank_svm)
```

## 11.6 Evaluting

The SVM model achieved an overall accuracy of 89.2%, which appears strong at first glance but is misleading given the class imbalance in the dataset, because if the model would classify all data points as "No", it would also achieve an accuracy of around 88%. The balanced accuracy of 0.567 — barely above the random guessing baseline of 0.5 — reveals the model's true weakness. While the model correctly identifies 99.1% of non-subscribers (Specificity), it only identifies 14.2% of actual subscribers (Sensitivity), meaning it misses approximately 86% of clients who would have agreed to a term deposit. This is reflected in the AUC of 0.726, indicating only fair discriminative ability between the two classes.

```{r evaluate-svm}
pred_svm <- factor(as.character(predict(bank_svm, bank_test)),
                   levels = c("No", "Yes"))

cm_svm <- confusionMatrix(pred_svm, bank_test_truth, positive = "Yes")

pred_svm_prob <- predict(bank_svm, bank_test, type = "prob")
roc_svm       <- roc(bank_test_truth, pred_svm_prob[, "Yes"])
auc_svm       <- as.numeric(auc(roc_svm))

svm_metrics <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", 
             "Balanced Accuracy", "AUC"),
  Value  = round(c(
    cm_svm$overall["Accuracy"],
    cm_svm$byClass["Sensitivity"],
    cm_svm$byClass["Specificity"],
    cm_svm$byClass["Balanced Accuracy"],
    auc_svm
  ), 3)
)

kable(svm_metrics, 
      caption = "SVM Model Performance Metrics",
      row.names = FALSE)

cm_plot_data <- as.data.frame(cm_svm$table) %>%
  mutate(
    label = case_when(
      Prediction == "No"  & Reference == "No"  ~ "True Negative",
      Prediction == "Yes" & Reference == "Yes" ~ "True Positive",
      Prediction == "Yes" & Reference == "No"  ~ "False Positive",
      Prediction == "No"  & Reference == "Yes" ~ "False Negative"
    )
  )


ggplot(cm_plot_data, aes(x = Reference, y = Prediction, fill = label)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(label, "\n n = ", Freq)),
            size = 5, fontface = "bold", color = "white") +
  scale_fill_manual(values = c(
    "True Positive"  = "#2ecc71",
    "True Negative"  = "#2ecc71",
    "False Positive" = "#e74c3c",
    "False Negative" = "#e74c3c"
  )) +
  labs(
    title    = "Confusion Matrix — SVM",
    x        = "Actual",
    y        = "Predicted"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r svm-save, include=FALSE}
saveRDS(bank_svm, "../models/svm_fit.rds")
cat("Saved: ../models/svm_fit.rds\n")
```

---

# 12. Neural Networks - Predicting subscribed

*Section lead: Aurelio Wyrsch*

Like SVM, neural networks can be used on classification tasks. By using the attributes as input nodes and the target variable **subscribed** as the output node, a neural network can learn complex, non-linear relationships between the features and the likelihood of subscription. This allows the bank to capture subtle patterns in the data that may indicate a client's propensity to subscribe, enabling more accurate predictions and better targeting of marketing efforts.

## 12.1 Training

Unlike for SVM, this time we don't use cross validation and hyperparameter tuning. Instead, we directly train a single model on the full training set. To address the class imbalance, we apply class weights during training, like we did for SVM, giving more importance to the minority class (subscribers) to help the model learn from those examples.

**Model:** nnet(subscribed ~ . , data = bank_train, size = 8, decay = 0.1, maxit = 500, weights = nnt_weights, trace = FALSE)

```{r train-nnt}
nnt_weights <- ifelse(bank_train$subscribed == "Yes", 7.5, 1)

set.seed(123)
bank_nnt <- nnet(subscribed ~ . , data = bank_train, size = 8, decay = 0.1, maxit = 500, weights = nnt_weights, trace = FALSE)
```

## 12.2 Evaluating

At a first glance, it looks like the neural network performed worse than the SVM, because the accuracy is now only 80.1%. Also the specificity is less with only 82.2% compared to 99.1% from SVM. However, the neural network outperforms the SVM across the other metrics, confirming that it is better suited for this classification task. The balanced accuracy of 0.734 represents a substantial improvement over the SVM's 0.567, indicating the model is now meaningfully distinguishing between subscribers and non-subscribers rather than defaulting to the majority class. As mentioned, the specificity of the model correctly identifies 82.2% of non-subscribers, while the sensitivity now successfully catches 64.6% of actual subscribers — a dramatic improvement over the SVM's 14.2%.

```{r evaluate-nnt}
pred_nnt <- factor(as.character(predict(bank_nnt, bank_test, type = "class")),
                   levels = c("No", "Yes"))

cm_nnt <- confusionMatrix(pred_nnt, bank_test_truth, positive = "Yes")

pred_nnt_prob <- predict(bank_nnt, bank_test, type = "raw")
roc_nnt       <- roc(bank_test_truth, pred_nnt_prob[, 1])
auc_nnt       <- as.numeric(auc(roc_nnt))

nnt_metrics <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity",
             "Balanced Accuracy", "AUC"),
  Value  = round(c(
    cm_nnt$overall["Accuracy"],
    cm_nnt$byClass["Sensitivity"],
    cm_nnt$byClass["Specificity"],
    cm_nnt$byClass["Balanced Accuracy"],
    auc_nnt
  ), 3)
)

kable(nnt_metrics,
      caption   = "Neural Network Model Performance Metrics",
      row.names = FALSE)

cm_plot_data_nnt <- as.data.frame(cm_nnt$table) %>%
  mutate(
    label = case_when(
      Prediction == "No"  & Reference == "No"  ~ "True Negative",
      Prediction == "Yes" & Reference == "Yes" ~ "True Positive",
      Prediction == "Yes" & Reference == "No"  ~ "False Positive",
      Prediction == "No"  & Reference == "Yes" ~ "False Negative"
    )
  )

ggplot(cm_plot_data_nnt, aes(x = Reference, y = Prediction, fill = label)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(label, "\n n = ", Freq)),
            size = 5, fontface = "bold", color = "white") +
  scale_fill_manual(values = c(
    "True Positive"  = "#2ecc71",
    "True Negative"  = "#2ecc71",
    "False Positive" = "#e74c3c",
    "False Negative" = "#e74c3c"
  )) +
  labs(
    title    = "Confusion Matrix — Neural Network",
    x        = "Actual",
    y        = "Predicted"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

```{r nnt-save, include=FALSE}
saveRDS(bank_nnt, "../models/nn_fit.rds")
cat("Saved: ../models/nn_fit.rds\n")
```
---

# 13. AI Usage

### Hanieh Jebeli

AI mainly helped with structuring the data cleaning process and speeding up coding tasks such as factor recoding and mutate() operations. It was also useful for understanding concepts like the offset term in the Poisson GLM, which I then verified using the course materials.
Some suggestions required manual checking. For example, AI initially recommended a log transformation of balance without considering negative values, so I adjusted the approach by creating a shifted balance variable.


### Enerel Khuyag

AI was mainly used to support coding, debugging, and interpretation throughout the Binomial GLM and GAM sections. It helped speed up repetitive R tasks such as generating evaluation metrics, plotting ROC curves, formatting model outputs, and troubleshooting modelling or package-related errors. AI was also used to better understand the theoretical differences between GLMs and GAMs and to improve the clarity of model interpretations and explanations.


### Aurelio Wyrsch

AI was primarily used to speed up repetitive coding tasks such as writing visualisations and evaluation chunks consistently across model sections. It was also helpful for answering small implementation questions. Every AI-generated solution was manually reviewed and tested before being included in the report. This was done because some suggestions were incorrect or did not fit the specific context of the data.

---

# 14. Conclusion

The **Linear Model** showed that job type, education, and loan status are significantly associated with balance. Management and tertiary educated clients hold substantially higher balances. These are financially stronger profiles and more likely to have capacity to invest.

The **Poisson GLM** revealed a U-shaped age effect: clients under 30 and above 60 subscribe at the highest rates. Education amplifies this effect tertiary-educated clients across all age groups subscribe more. Housing loans consistently reduce subscription rates, suggesting that existing financial commitments are a key barrier.

Together, these two models suggest that the bank should prioritise young and older clients with higher education and no housing loan and be cautious about over-contacting middle-aged clients who are the least responsive segment.

All four classification models were evaluated on the same held-out test set. The Binomial GLM and GAM performed similarly, achieving high accuracy (89%) but very low sensitivity (17–18%), meaning they correctly identified only a small fraction of actual subscribers. This is a direct consequence of the class imbalance — without any adjustment, both models default to predicting "No" for the vast majority of clients. The SVM, despite using class weights, showed similarly poor sensitivity (14.2%) and the lowest AUC (0.729), likely due to being trained on only 30% of the training data for computational reasons.

The Neural Network was the strongest performer overall. By applying class weights during training, it achieved a sensitivity of 64.6% and a balanced accuracy of 0.734 — substantially better than the other three models at identifying actual subscribers. Its AUC of 0.789 also edged out the GAM (0.780) and GLM (0.775). The trade-off is a drop in overall accuracy to 80.1%.

From a business perspective, the Neural Network can clearly improve targeting new subscribers. Of the 2,105 clients the model flagged as likely subscribers, 683 were true positives (actual subscribers) and 1,422 were false positives (non-subscribers incorrectly identified as likely subscribers), yielding a precision of approximately 32%. While this may seem low, it still represents a meaningful improvement over random targeting. In the original dataset, only 1 in 7 clients (12%) actually subscribed — by focusing exclusively on the clients flagged by the model, the bank would expect a successful call roughly every 3 attempts, more than doubling the baseline conversion rate. This means the model, despite its limitations, can still add practical value by helping the bank allocate its marketing resources more efficiently and reduce the number of unproductive calls.







